\bf Model Visualization Techniques for a Social Network Model

### 2018-10-01 paper-child.Rnw
revised abstract, finished revising intro section. Quite pleased with it

\section{Introduction \& Background}
% peices of old intro
%Social networks, such as collaboration networks between academic researchers or friendship networks 
% new intro 
%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues. These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals. There are even statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}. These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes. \textit{Dynamic social networks} are more realistic because they contain multiple observations of the same network over time, and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms. Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our world.

\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}. These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree. 
%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example 
With these models, we can examine our assumptions that people with common interests are more likely to become friends. For example, we may assume that high schoolers who drink heavily are more likely to be friends with others who drink heavily. SAOMs allow us to incorporate this behavioral information on shared interests into the modeling process.

\par Unlike other network models, SAOMs are not analytically tractable. Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models. %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. 
In order to estimate the parameters of a SAOM, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user. In order to better understand the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results. Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.

In the remainder of this introduction, we first describe the structure of network data and different network visualization techniques. Then, we carefully define the many moving pieces in SAOMs. Finally, we briefly ouline the SAOM fitting process to a set of dynamic network data.

@@

\noindent \textbf{Visualizing Network Data:} Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have an inherent structure.

@@

There are two primary methods used to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}. For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$. This data is shown using the two visualization methods in Figure~\ref{fig:toyplots}.

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}. Because there is usually no natural placement of the points, a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}. Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. %Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. 
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper will use the KK layout.

\par The second method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\begin{equation}\label{eq:adjmat}
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\end{equation}
In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row and column representing a node. The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.

\par Each visualization comes with its own advantages and disadvantages. For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness or linetype. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, because this type of visualization features the edges. Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.% Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization. This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. 
Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.

@@

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: actor, tie, and temporal information. It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates. The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways. Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.

@@

We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$. We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we will denote by $X(T)$. This CTMC is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and currently reachable states of the network. \\

\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ gets an opportunity to changes its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have just one set of rate parameters in the overall model. In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that if $j = i$, no change is made.) A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$, as shown in Equation~\ref{eq:c1}.

\begin{equation} \label{eq:c1}
 \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}

From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed with expected value $(n\alpha_m)^{-1}$, where $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter in \texttt{RSiena} is simple: the method of moments is used to estimate the rate with the statistic 
\begin{equation}\label{eq:cdiff}
C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. \\

\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC. The node $i$, which is given the opportunity to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change. For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x, \mathbf{Z}) = \sum_{k=1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates, and $K$ the total number of parameters in the objective function. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if all changes will result in lower values of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$.

According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes. (These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given inTable~\ref{tab:effects}. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model. The formulas for the corresponding statistics for all effects are provided in \citet{RSienamanual}.

The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters. The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are \emph{only} functions of the current network state, $x$. They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$. Some possible covariate parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures.}

@@

The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. Let $x(i \leadsto j)$ be the network identical to $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$. The probility that the tie $x_{ij}$ changes is defined as:

\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
When the value of the objective function is high at the current state, the probability of not making a change is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM. For more on CTMCs, see \citet{ctmc}. We explore these probabilities further in Section~\ref{sec:algorithms}

% \subsubsection{Continuous Time Markov Chain (CTMC)}
% \st{
% In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states for a directed network with $n$ nodes. We denote the state space as $\mathcal{X}$, a set which contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and there are $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
% 
% The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states. The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.  Then the $bc^{th}$ entry of $Q$ is:
% \[ q_{bc} = \begin{cases}
%      q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
%       0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
%       -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
%    \end{cases}
% \]
% 
% Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM. %Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
% 
% %XXX HH can you help? I need a way better transition than that XXX
% }

>>>>

\section{Introduction \& Background}
% peices of old intro
%Social networks, such as collaboration networks between academic researchers or friendship networks 
% new intro 
%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues. These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals. There are even statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}. These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes. \textit{Dynamic social networks} are more realistic because they contain multiple observations of the same network over time, and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms. Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.

\par One set of models for dynamic social networks are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}. These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree. 
%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example 
With these models, we can examine our assumptions that people with common interests are more likely to become friends. For example, we may assume that high schoolers who drink heavily are more likely to be friends with others who drink heavily. SAOMs allow us to incorporate this behavioral information on shared interests into the modeling process.

\par Unlike other network models, SAOMs are not analytically tractable. Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models. %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. 
For SAOM parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user. To better understand SAOMs and their fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results. Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.

In the remainder of this section, we introduce the structure of network data and two different network visualization (net-vis) techniques. Then, we carefully define the many moving pieces that make up a SAOM. Finally, we briefly ouline the SAOM fitting process that we use throughout this paper.

@@

\noindent \textbf{Visualizing Network Data:} Network visualization, also called graph drawing, is a prominent subfield of network analysis \citep{graphdrawing}. Visualizing network data is inherently challenging because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have such built-in structure.

@@

The lack of a native coordinate system has led to two very different methods to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}. We demonstrate these with a simpe example. Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}. Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with one of the many layout algorithms available \citep{drawingalgs}. Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. %Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. 
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper were drawn with the KK layout algorithm.

\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as:
\begin{equation}\label{eq:adjmat}
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j, \quad i \neq j \\
                                   0 & otherwise
  \end{cases}
\end{equation}
%In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. 
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row and column representing a node. The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.

\par Each net-vis method has its own advantages and disadvantages. For instance, \emph{paths} between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness or linetype. Visualizing a node-level variable into an adjacency matrix visualization is not as simple, because this type of visualization features the edges. Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}. % Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization. This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. 
Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore our example data and SAOMs.

@@

Stochastic Actor-Oriented Models (SAOM) are unique because they model \textit{dynamic} networks. We use SAOMs to study the change of a network in time while also examining the structural and node covariate efects. The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear. Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.

@@

A \textit{dynamic network} is a network, that consists of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$. We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain, (CTMC), which we will denote by $X(T)$. The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e.$x(t_1) \equiv X(0)$ and $x(t_M) \equiv X(\infty)$. Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's \textit{rate function}, where one actor at a time is ``selected" to make a change. Once selected, the actor tries to maximize its \textit{objective function} based on the current state and possible future states of the network. \\

\noindent \textbf{The Rate Function} In general, the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters,   $\boldsymbol{\alpha}$. With this general formulation, it is possible that each node will have a different rate of change. Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is a constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m-1})$ to $x(t_m)$ for $m = 2, \dots, M$. The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$. A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ follows an exponential distribution:
\begin{equation} \label{eq:c1}
 \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}

\noindent From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed, and has expected value $(n\alpha_m)^{-1}$. The estimation of $\alpha_m$ in \texttt{RSiena} is simple: the method of moments (MoM) is used to estimate the rate with the statistic 
\begin{equation}\label{eq:cdiff}
C_m = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. \\

\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we consider the objective function for each node separately. Suppose node $i$ is selected to change at the next step in the CTMC. This node, called the \textit{ego} node, has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and will not change any of their ties at the next step in the CTMC. The ego node $i$ in the current network state $x$, will make a change that maximizes its \emph{objective function}, written as:
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x, \mathbf{Z}) = \sum_{k=1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for the network $x$, covariates $\mathbf{Z}$, and $K$ the total number of parameters in the objective function. The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To maximize $f_i$, $i$ will either destroy a tie where $x_{ij} = 1$, create a tie where $x_{ij}=0$, or make no change if all changes will result in lower $f_i$.

The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters. The structural parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$. According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes. The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Table~\ref{tab:effects}. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model. The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two selected covariate parameters and corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.

@@

With the objective function, we can define the probability that the ego node $i$ chooses to change its tie to alter node $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. Let $x(i \leadsto j)$ be the network identical to $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$. The probility that the tie $x_{ij}$ changes is defined as:
\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
%When the value of the objective function is high at the current state, the probability of not making a change is also high.
In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. We explore these probabilities further in Section~\ref{sec:algorithms}. The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change. For more on CTMCs, see \citet{ctmc}. 
% \subsubsection{Continuous Time Markov Chain (CTMC)}
% \st{
% In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states for a directed network with $n$ nodes. We denote the state space as $\mathcal{X}$, a set which contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and there are $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
% 
% The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states. The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.  Then the $bc^{th}$ entry of $Q$ is:
% \[ q_{bc} = \begin{cases}
%      q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
%       0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
%       -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
%    \end{cases}
% \]
% 
% Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM. %Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
% 
% %XXX HH can you help? I need a way better transition than that XXX
% }

### 2018-10-01 paper.Rnw
fix a fig ref
revised abstract, finished revising intro section. Quite pleased with it

\bigskip
\begin{abstract}
Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks. Unlike other network models, SAOMs are not very well understood. We use model visualization techniques introduced in Wickham et al (2015) in order to gain a better understanding of the models and their behavior. Viewing the model in the data space places these models into the field of network drawing, while exploring collections of models helps with parameter interpretation. Finally, delving into the fitting algorithms provides more insight into the underlying mechanisms. With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of SAOMs for social network analysts.
\end{abstract}

>>>>

\bigskip
\begin{abstract}
Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks. Unlike other network models, SAOMs are analytically intractable, and therefore hard to interpret. We use model visualization techniques introduced in Wickham et al (2015) better understand these models and their behavior. We use a three-pronged attack: (a) a network visualization to view the \emph{model}, (b) exploring collections of models for parameter interpretation, and (c) delving into the fitting algorithms to gain insight into the underlying mechanisms. With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of SAOMs for social network analysts.
\end{abstract}
